Nice people or potential cooperators when keeping promises

### 2019-08-14 behavioral-promises/behavioral-promises.Rmd
Update manuscript

abstract: |
  Promises increase cooperation between non-genetically related individuals, however, lies and betrayals occur daily in close social relationships. Despite this, the effect of social closeness on the decision to keep or break a promise has not been studied. We conducted an experiment in which subjects could freely decide if they broke or kept their promise to three partners with different levels of social closeness: null (computer), low (strange) and high (friend). If subjects keep their word with the computer, it provides evidence in favor of intrinsic motivation (comply to do the correct thing), otherwise extrinsic motivation would be supported (comply to facilitate future cooperation). Using Hierarchical Bayesian Modeling, we find that if social closeness increases, the probability of breaking promises decreases monotonically, the evidence ratio in favor to the hypothesis stranger < computer was 7.47, and friend < stranger was 8.64. Furthermore, using Bayesian Cumulative Modeling we find that subjects were very consistent between their promises and the subsequent decisions they make. Although is more probable that our subjects keep their promises to any of the three partners, there is a proportion of transgressions to the computer, whose inference excludes zero with a 95% posterior probability. Also, subjects never broke the promise to their friends. Our results indicate that subjects are predominantly "nice people" because they keep their promises even to partners without social closeness, however, if the subjects are not certain that their partner will also be a "potential cooperator", dishonesty emerges in the form of broken promises.

@@

Human are social animals and we get survival advantages from being such. Historically, belonging to a social group has provided protection and guaranteed access to resources such as food, and currently, there has been evidence that belonging to a social group has a positive effect on longevity, physical and mental health [@Tough2017; @Holt-Lunstad2018]. The relationship between social closeness and cooperation is key to the formation of large groups of individuals without a genetic relationship, and such groups form the basis of communities, societies, and nations, arguably constituting one of the most fundamental conditions for human survival [@Fehr2003; @Fehr2018]. An important requirement for cooperation between individuals is communication. A meta-analysis of 45 studies reported a large positive effect ($d =  1.01$) of communication in cooperation regardless of the communication medium [@Balliet2010]. In parallel, it is known that after communication sessions between completely unknown individuals there is an increase in subjective social closeness [@Aron1997]. Thus, these findings suggest a positive relationship between communication, cooperation, and social closeness.

One of the forms of communication that have received the most attention in psychological game theory is "the promise". In an interaction between sender and recipient, it is believed that promises influence beliefs of the recipient, generating trust and cooperation [@Charness2006; @Vanberg2008]. However, there are situations where trust is betrayed, promises are broken or people deceived. The General Social Survey (GSS) in 2018 showed that approximately 11% of people surveyed responded that they have had sex with someone different to their partner while they are married [@GSS2018]. In economic and social terms, the United Nations Organization has stressed that deception (corruption) has a global cost of at least 5 percent of the annual gross domestic product (GDP), and it is a generator of conflict and instability in all nations [@UN2018].

Despite the widespread presence of interactions between human belonging to the same group, it is not known how social closeness between participants can affect keeping or breaking promises. This is important because it has been pointed out that lies and betrayals of trust occur daily in close relationships, such as friends, workmates, or classmates. And, their consequences can be serious such as loss of the relationship, work or damage to reputation [@DePaulo2004]. In the laboratory, these transgressions to trust have been explicitly studied in experiments where subjects have incentives to lie or not keep their promises. However, the vast majority of these studies have been conducted in people who do NOT know each other [@Gneezy2005; @Gneezy2013; @Baumgartner2009; @Baumgartner2013; @Charness2006; @Mazar2008; @Fischbacher2013a]. Some studies address deception in close interpersonal relationships using self-report measurements, however, they have limitations due to social desirability bias [@DePaulo1998; @DePaulo2004] or rather explore the development of deception detection skills in same-sex friends [@Anderson2002].

In this study, we wanted to understand the effects of three partners with different levels of social closeness on keeping or breaking promises, as well as on cooperation, using an experimental trust game. We analyzed it in light of the two main motivations that have been pointed out in the literature regarding keeping promises which are: instrumental and intrinsic [@Baumgartner2009]:

1. The _instrumental_ suggests that promises are kept to make future cooperation easier.

2. The _intrinsic_ mentions that the promises are kept to do what is morally right.

In our study, subjects perform a standard trust game in pairs with three phases: _first_, the trustee makes a promise to pay half of his earnings regardless of who his investor is; _second_, the investor receives the promise and decides if he invests his initial budget; _third_, in case the investor has given the budget, the trustee faces the decision to pay or not to pay half of his earnings. To evaluate the effect of social closeness, our subjects participate in the role of trustee in front of three partners with different levels of social closeness (null, low and high) in the role of the investor: a computer, a stranger, and a friend. The manipulations mentioned allow us to evaluate several hypotheses, the first one was proposed a priori and is derived from a larger research project (you can check the preregistration of the hypothesis here: https://osf.io/u97fd), while the following are exploratory:

1. Social closeness will reduce the decision of breaking the promise. According to the _instrumental_ motivation, we expect that subjects keep their promises with friends for the purpose of facilitating future cooperation, which can be extended even beyond the trials in the experiment. We also expect, although to a lesser extent, that the subjects keep promises to strangers, with the purpose of facilitating cooperation at least during the trials during the experiment. Finally, we anticipate that the participants break the promises to the computer because it is a partner without social closeness and they could not ensure cooperation in future trials. It should be noted that if the participants keep the promises to the computer, evidence would be given in favor of _intrinsic_ motivation.

2. There will be an effect of social closeness on cooperation, regardless of promises. We expect that there is more probability of paying the friend than the stranger and more probability of paying the stranger than the computer. According to the _instrumental motivation_, the cooperation will be greater for partners with whom the subject anticipates greater cooperation in the future (friend > stranger > computer).

3. Finally, we will obtain two subsamples according to the cooperation rate of the subjects as was done in another study [@Baumgartner2009]. However, what in the aforementioned study was classified as a group of "honest" and "dishonest", we will show that in our sample it is not supported because if the groups differ in their payment rate it will be due to the difference in their level of commitment expressed by promises.

@@

We recruited 45 subjects (15 male) from the National Autonomous University of Mexico, the age range from 19 to 33 years and their minimum educational level were bachelor. Subjects were asked to come to the study with a “close” friend, who fulfilled the following characteristics: 1) he/she was matched by sex, 2) did not have a family bond and 3) was not a person with a sentimental or sexual relationship. From this sample, 30 subjects (15 men) performed the task in a magnetic resonance imaging (MRI) scanner, however, their image data was not analyzed for the present work.

@@

All subjects performed an adaptation of the trust game with promises, using hypothetical monetary rewards [@Baumgartner2009]. The variation with respect to the original task is that in our experiment three partners were presented with three levels of social closeness in the role of the investor: computer (no closeness), strange (low closeness) and friend (high closeness), while the participant acted as the trustee. The task was programmed in PsychoPy2 version 1.84.2 [@Peirce2019; @Peirce2008] and consisted of 24 trials between two players: trustee and investor. The trustee originally has 0 Mexican pesos and the investor has 2 pesos, the investor is presented with the opportunity to give his money to the trustee or keep it. If investor gives his money, it is multiplied by 5 pesos, so that the trustee has 10 pesos. Finally, the trustee decides to pay half to the investor or keep the 10 pesos. The structure described is repeated in 24 trials, however, in 4 of the trials the trustee can send a promise to the investor, the promises were that always, mostly, sometimes, or never would pay back. Each promise was valid for three trials so that 12 of the trials have the effect of the promise and the other 12 do not. Each partner performs the role of investor in 8 of 24 total trials, however, both the promises and social closeness conditions were presented to the trustee in pseudorandom order.

Since our main interest was the trustee’s behavior, investors’ decisions were programmed _a priori_ to give their amount in 6 trials and in 2 did not randomly. The covert story for all our subjects was that they would be playing in real-time with their friend (whom they brought to the study), the stranger (was told he would be another same-sex unknown person) and the computer. After the study ended, the participants were debriefed about our own deception. A diagram of the chronology of the experimental task with the duration of each phase in seconds is shown in Figure 1.

@@

Fixation phase consisted only of a period in which the subject paid attention without performing any particular behavior. Then, **promises phase** in part A1 indicated that in the following three trials they could decide without the effect of the promise. In A2, the subject was asked to decide between 1) always, 2) mostly, 3) sometimes or 4) never pay back. This pattern was repeated. In the **anticipation/assignation phase**, subjects were told who their partner was for that trial (computer, stranger or friend) and was given the message that their partner was making their decision. Later in the **decision phase**, the subject was informed if his partner invested his $2 or not, he was reminded of his promise level (if they were trails with promises) and, in case the partner had invested, he was asked to decide whether to pay back or not. Finally, payments for each trial were shown in the **feedback phase** and the sequence was repeated until the task finished.

@@

```{r fig2, fig.cap='Pay back proportion by partners and promises levels'}

@@

```{r fig3, fig.cap='Posterior probability of Promise Breaking'}
p3

@@

```{r fig4, fig.cap='Posterior estimates in logit scale'}

@@

```{r fig5, fig.cap='Posterior predictive over pay back rate', out.width='100%', out.height='42%'}

@@

```{r fig6, fig.cap='Posterior predictive over pay back rate, Social Closeness Varying effects by Promise levels'}

@@

```{r fig7, fig.cap='Promise choices by group'}

@@

To our knowledge, this is the first experimental study that incorporates social closeness as a predictor of the decision to break promises with socially relevant partners. In previous studies, the participants remain anonymous during the course of the tasks [@Baumgartner2009], their measurements to evaluate transgressions are self-reported [@DePaulo1998; @DePaulo2004], they do not directly quantify the breaking of promises [@Vanberg2008; @Charness2006], or as we will discuss later, they use heterogeneous measures of social closeness [@Glaeser2000]. Our experiment is the first one to include a partner without social closeness (computer). Our data suggest that the mere fact of considering that subjects play with a human diminishes the probability of breaking the promise. It is more likely to break the promise to the computer than to the stranger, even though that partner was not known. Also, null social closeness is important because it allows us to explore the intrinsic motivation to keep the promises. If humans keep their word because it is morally correct, we would anticipate that the probability of breaking a promise was very low in the three partners and, particularly, the inference regarding the probability of breaking the promise to the computer would include zero. However, we found that, although low, the probability was greater than zero, which provides evidence that seems to contradict, at least in degree, the hypothesis of intrinsic motivation. We cannot exclude the possibility that the subjects keep promises mainly due to moral motivation because, even in the case of the computer, the subjects kept a large proportion. We can, however, exclude that this is their only motivation.

On the other hand, high social closeness allows us to explore the instrumental motivation to keep promises [@Baumgartner2009]. In our study, it was more likely to break the promise to the stranger than to the friend. If humans keep their word to facilitate cooperation in the future, we would anticipate that with a partner of high social closeness (with whom they are very likely to cooperate, even after the experiment) there would be less of a chance to break the promise. In fact, the promise to friends was not broken on any occasion, hence the inference in this case, was met.

The findings concerning motivations could be explained by the social norm called "conditional cooperation", which indicates that the belief that other people cooperate at high levels also induces high levels of cooperation [@Fehr2018]. Thus, the lack of certainty regarding the decisions that the computer would make could explain the probability of breaking the promise that exists towards this partner. Similarly, the information that the subjects have regarding as their friend's behavior -even before the experiment- could explain the high levels of cooperation and keeping of promises towards this partner.

Cooperation has been studied with diverse experimental manipulations. In one study, the contribution in a public goods game was evaluated during several trials in a group made up of the same individuals (partners), and compared to another group of new different subjects for each trial (strangers) [@Croson1996]. In the condition of partners the cooperation was greater because it is a stable group compared to the group of strangers. The mentioned study does not give details regarding the recruitment of participants, so we could assume that even in the condition of partners these are individuals who do not consider themselves socially close.

Another study showed an increase in cooperation in a trust game when individuals were socially close [@Glaeser2000]. In this study, subjects knew each other and the researchers carefully measured several variables regarding their social connection.However, some individuals who arrived together were allowed to perform the task between themselves and others were paired with unclear criteria. The above allows social closeness between partners to be heterogeneous, allows for romantic relationships and assumes that two individuals who come together to class are considered close to each other. Also, since it is a study of one-shot, it excludes the possibility of evaluating how the same subject varies his behavior based on different levels of social closeness.

To try to homogenize social closeness between our subjects and their partners, they performed the task with a same-sex friend considered to be close, emphasizing their companion could not be a romantic or sexual partner. Bayesian analysis of the IOS scale showed evidence in favor that the degree of social closeness between 30 of our 45 subjects and their friends was the same. Results of the two mentioned studies also found that social closeness enhance cooperation [@Croson1996; @Glaeser2000], in the future would be valuable to homogenize both the performed tasks and the statistical procedure used, to establish similarity in effects magnitudes.

Our methodological contribution to the field is the use of Bayesian inference tools: Hierarchical Models and Cumulative Models. The Hierarchical Models, allowed us to model the effect of the experimental conditions on the individual response. These models assume that lower level observations (e.g. decisions of each individual) are nested in higher level units (e.g. individual subjects). Within-subjects designs have traditionally been analyzed with repeated measures AN(C)OVA, however, Hierarchical Models grant the advantages of naturally dealing with unbalanced data, include categorical and/or numerical predictors, explicitly incorporate individual variability, among others [@Vuorre2018; @Gelman2006; @McElreath2018; @Burkner2017].

In our case, the Hierarchical Model allowed us to capture how the experimental conditions affected the decisions of each subject and the differences between them. For example, there are obviously motivated instrumental subjects such as number 8 or 12 who are very sensitive to the identity of their investors and modify their cooperation under social closeness. At the same time, there were notably intrinsic subjects such as 9 and 10 who cooperate all the time regardless of who their partners are. The Hierarchical Model naturally includes this information for the estimation of population effects, which, if not considered, would lead to inaccurate inferences[@Burkner2017; @Gelman2006; @McElreath2018].

On the other hand, the Cumulative Model allowed us to capture the strength of the commitment expressed through promises. A very frequent problem that has been pointed out recently is that analyzing ordinal data with methods that assume that observations are metric can lead to serious inference errors [@Liddell2018]. The Cumulative Models assumes that the observed responses come from the categorization of a continuous latent variable [@Burkner2019]. In our study, we could identify that groups with different payment proportions, which in other studies have been called "dishonest" and "honest"[@Baumgartner2009; @Baumgartner2013], differ in the commitment they express through their promises. Considering that the tools to perform the Cumulative Models are relatively recent, in future studies, it would be valuable to also use the new alternatives for inference.

Although the Hierarchical and Cumulative models are not tools exclusively for Bayesian inference, their application from this approach represents several advantages compared to frequentist statistics. Classic problems such as multiple comparisons, the decision of when to stop collecting subjects (stopping rule), or use of planned comparisons versus post hoc, are not factors that affect the Bayesian approach [@Dienes2011]. In Bayesian inference, hypothesis testing makes formal use of probability to express the plausibility of theories, and in our case, we were able to obtain evidence ratios regarding the extent of our data support hypotheses, regardless if these were proposed _a priori_ or _post hoc_.

@@

A limitation of this paper is the use of hypothetical rewards compared to real rewards. A relatively recent study on decision-making reported that there is more loss aversion when subjects have real monetary rewards compared to hypothetical in a risk task [@Xu2016]. However, other works do not report differences between the use of both types of rewards in self-control, temporal and social discount tasks [@Locey2011; @Johnson2006; @Rachlin2006]. Likewise, it can be argued that our results have theoretical congruence [@Fehr2018] and are in the same direction as other works that use real money [@Croson1996; @Glaeser2000], thus, there are not many reasons to expect that other types of rewards would modify our results. Another possible limitation is the use of multiple trials with each partner. Although the decision to use repeated measures during our design serves more the purpose of reducing contaminant sources and increasing internal validity [@Maxwell2018], to have more clarity regarding the difference in the intrinsic and instrumental motivations, the studies could benefit from only include one trial for each partner. Play a one-shot with each partner avoids the possibility that multiple trials could generate the belief that the computer can also vary its behavior according to the decisions of the trustee. Finally, subjects that performed the study inside the MRI scanner may differ in their responses than subjects outside of the scanner. However, our analysis did not show this explicitly.

@@

Subjects seem to keep their promises by a combination of _intrinsic_ and _instrumental_ motivations. They are predominantly "nice people" because they keep their word even with investors with no social closeness. However, their trustworthiness is far from perfect, since there is a small proportion of betrayals committed towards this partner. It seems that if subjects cannot be sure their investor will be a "potential cooperator", dishonesty may emerge in the form of broken promises. Framing the finding in terms of the theoretical predictions of the social norm "conditional cooperation", social closeness decreases the probability of breaking the promise and increases cooperation, and suggests the correspondence between commitment expressed through promises and subsequent behavior.

>>>>

abstract: |
  Promises increase cooperation between non-genetically related individuals, whereas lies and betrayals occur daily even in close social relationships. Despite this, the effect of social closeness on the decision to keep or break a promise has not been thoroughly studied. We conducted an experiment in which subjects could freely decide if they broke or kept a promise to three partners with different levels of social closeness: null (computer), low (strange) and high (friend). If subjects kept their word with the computer, it provides evidence in favor of intrinsic motivation (comply to do the correct thing), while extrinsic motivation would be supported (comply to facilitate future cooperation). Using Hierarchical Bayesian Modeling, we found that as social closeness increases, the probability of breaking promises decreases monotonically. The evidence ratio in favor of the hypothesis stranger < computer was 7.47, and friend < stranger was 8.64. Furthermore, using Bayesian Cumulative Modeling we found that subjects were very consistent between their promises and their choices. Although it was more probable that our subjects kept their promises to any of the three partners, there was a proportion of transgressions to the computer, whose inference excludes zero with a 95% posterior probability. Also, our subjects never broke the promise to their friends. Our results suggest that subjects were predominantly "nice people" because they kept their promises even to partners without social closeness. However, if the subjects were not certain that their partner will also be a "potential cooperator", dishonesty emerged in the form of broken promises.

@@

Humans are social animals and we obtain survival advantages from proper socialization. Historically, belonging to a social group has provided protection and guaranteed access to resources such as food, and currently, there is evidence that belonging to a social group has a positive effect on longevity, physical and mental health [@Tough2017; @Holt-Lunstad2018]. The interaction between social closeness and cooperation seems to be a key feature to the formation of large groups of individuals without a genetic relationship, and such groups form the basis of communities, societies, and nations, arguably constituting one of the most fundamental conditions for human survival [@Fehr2003; @Fehr2018]. A recent meta-analysis showed that communication, regardless of the medium, is the most important element for cooperation between individuals [@Balliet2010]. In parallel, there is evidence that following communication sessions between completely unknown individuals, there is an increase in subjective social closeness[@Aron1997]. These findings suggest the importance of the relationship between communication, cooperation, and social closeness.

One of the forms of communication that have received the most attention in psychological game theory is “the promise”. In an interaction between a sender and a recipient, it is believed that the sender's promises influence the recipient’s beliefs, influencing their ability to trust and cooperate [@Charness2006; @Vanberg2008]. Commonly, however, there are situations where promises are broken, trust is betrayed and people are deceived. The General Social Survey (GSS) in 2018 showed that approximately 11% of people surveyed responded that they have had sexual intercourse with someone different to their partner while they were married  [@GSS2018]. In economic and social terms, the United Nations Organization has stressed that deception (corruption) has a global cost of at least 5 percent of the annual gross domestic product (GDP), and it is a generator of conflict and instability in all nations  [@UN2018]. Hence, breaking promises may greatly affect our own social groups and society itself.

Despite the widespread presence of interactions between human belonging to the same group, it is not known how social closeness between participants can affect keeping or breaking promises. This is important because it has been pointed out that lies and betrayals of trust occur daily even in close relationships such as friends, coworkers, or classmates, and their consequences can be serious, from loss of the relationship to work loss and damage to reputation [@DePaulo2004]. The two main motivations that have been pointed out in the literature regarding keeping promises are: instrumental and intrinsic [@Baumgartner2009]. The instrumental suggests that promises are kept to make future cooperation easier while the intrinsic suggests that the promises are kept to do what is "morally right". In the laboratory, the betrayals of trust have been explicitly studied in experiments where subjects have incentives to lie and break promises. However, the vast majority of studies have been conducted in people without any social closeness (strangers) [@Gneezy2005; @Gneezy2013; @Baumgartner2009; @Baumgartner2013; @Charness2006; @Mazar2008; @Fischbacher2013a]. Some studies have addressed deception in close interpersonal relationships using self-report measurements or rather explore the development of deception detection skills in same-sex friends, with clear limitations due to the social desirability bias [@DePaulo1998; @DePaulo2004; @Anderson2002].

In this study, we wanted to understand the effects of social closeness on keeping or breaking promises and cooperation, by means of an experimental trust game with three partners with different levels of social closeness. For this, subjects performed a standard trust game in pairs with three phases: 1) the trustee makes a promise to pay half of his earnings regardless of who his/her investor; 2) the investor receives the promise and decides if he/she invests his/her initial budget; 3) in case the investor has given the budget, the trustee faces the decision to pay or not to pay half of his earnings (keep or break promise). To evaluate the effect of social closeness, our subjects participate in the role of trustee with three partners with different levels of social closeness (zero, low and high) in the role of the investor: a computer, a stranger, and a close friend. Our main pre-registered hypothesis was:

1. Social closeness will reduce the decision of breaking the promise. According to the instrumental motivation, we expect that subjects keep their promises with friends for the purpose of facilitating future cooperation, which probably can be extended even beyond the trials in the experiment. We also expect, although to a lesser extent, that the subjects keep promises to strangers, with the purpose of facilitating cooperation during the experiment. Finally, we anticipate that the participants break the promises to the computer because it is a partner without any social closeness and they could not ensure cooperation in future trials. It should be noted that if the participants keep the promises to the computer, evidence would be given in favor of intrinsic motivation.

Our secondary exploratory hypotheses were:

2. There will be an effect of social closeness on cooperation, regardless of promises. According to the instrumental motivation, the payment will be greater for partners with whom the subject anticipates greater cooperation in the future (friend > stranger > computer).

3. We hypothesize that the subjects will cooperate according to their degree of commitment expressed through promises.

@@

We recruited 45 subjects (15 male) from the National Autonomous University of Mexico, age range from 19 to 33 years, their minimum educational level were bachelor and right handed (for MRI). Subjects were asked to come to the study with a “close” friend, who fulfilled the following criteria: 1) he/she was matched by sex, 2) did not have a family bond and 3) was not a person with a sentimental or sexual relationship. From this sample (n = 45), 30 subjects (15 men) performed the task in a magnetic resonance imaging (MRI) scanner, however, their image data was not analyzed for the present work.

@@

All subjects performed an adaptation of the trust game with promises, using hypothetical monetary rewards [@Baumgartner2009]. The variation with respect to the original task was that, in our experiment, three partners with three levels of social closeness were presented in the role of the investor: computer (no closeness), strange (low closeness) and friend (high closeness), while the participant acted as the trustee. The task was programmed in PsychoPy2 version 1.84.2 [@Peirce2019; @Peirce2008] and consisted of 24 trials between two players: trustee and investor. The trustee originally had 0 Mexican pesos and the investor had 2 pesos, the investor was presented with the opportunity to give his/her money to the trustee or keep it. If the investor gave his/her money, it was multiplied by 5 pesos, so that the trustee had 10 pesos. Finally, the trustee had to decide if he/she wanted to pay half to the investor or keep the 10 pesos. The structure described was repeated in 24 trials, however, in 4 of the trials the trustee could send a promise to the investor, the promises were that he/she would: always, mostly, sometimes, or never pay back. Each promise was valid for three trials, so that 12 of the trials had the effect of the promise and the other 12 did not. Each partner performed the role of investor in 8 of 24 total trials, however, both the promises and social closeness conditions were presented to the trustee in pseudorandom order.

Since our main interest was the trustee’s behavior and to avoid unbalanced responses from investors, the investor was always the computer although the trustee was shown to have 3 types of investors, and the investors’ decisions were programmed _a priori_ to randomly give their amount in 6 trials and withhold the amount in 2 trials. The covert story for all our subjects was that they would be playing in real-time with their friend (whom they brought to the study), the stranger (was told it would be another same-sex unknown person) and the computer. After the study ended, the participants were debriefed about our deception. A diagram of the the experimental task is shown in Figure 1.

@@

Promise phase in part A1 indicated that in the following three trials they could decide without the effect of the promise. In A2, the subject was asked to decide between 1) always, 2) mostly, 3) sometimes or 4) never pay back. This pattern was repeated. In the anticipation/assignation phase, subjects were told who their partner was for that trial (computer, stranger or friend) and was given the message that their partner was making their decision. Later in the decision phase, the subject was informed if his partner invested his $2 or not, he was reminded of his promise level (if they were trails with promises) and, in case the partner had invested, he was asked to decide whether to pay back or not. Finally, payments for each trial were shown in the feedback phase and the sequence was repeated until the task finished.

@@

```{r fig2, fig.cap='Pay back proportion by partners and promises levels, the x-axis represents all decisions made to each partner, the different shades of blue shows the proportions in which the subjects paid back or not. The y-axis presents how these proportions vary according to the promises made.'}

@@

```{r fig3, fig.cap='Inference regarding the probability of breaking the promise depending on social closeness. The x-axis shows the posterior probability of breaking the promise, and the y-axis indicates the three partners. The dotted line shows the probability that breaking the promise would be zero.'}

p3

@@

```{r fig4, fig.cap='Posterior estimates in logit scale. We show posteriors distributions of the hierarchical Bayesian model coefficients for the effect of experimental conditions on the decision to cooperate. The x-axis shows the magnitude of the logit scale coefficients, and the y-axis shows the experimental conditions associated with that effect.'}

@@

```{r fig5, fig.cap='Posterior predictive over pay back rate. Correspondence between payment proportions and model predictions depending on the game partner for the first twelve subjects in the sample. The x-axis pointed the three playmates when the subjects had made a promise, while the y-axis indicates the proportion of decisions in which subjects payback.', out.width='100%', out.height='42%'}

@@

```{r fig6, fig.cap='Posterior predictive over pay back rate, social closeness varying effects by promise levels. Correspondence between payment proportions and model predictions depending on the game partner at the group level. The x-axis pointed the three playmates when the subjects had made a promise, while the y-axis indicates the proportion of decisions in which subjects payback.'}

@@

```{r fig7, fig.cap='Promise choices by group. We show posterior probability estimates regarding the level of promise depending on the group. The Low group (which paid 58/100 of the trials) is more likely to promise mostly, while the High group (which paid 83/100 of the trials) is more likely to promise always.'}

@@

To our knowledge, this is the first experimental study that incorporates social closeness as a predictor of the decision to break promises with socially relevant partners. In previous studies, the participants remain anonymous during the course of the tasks [@Baumgartner2009], their measurements to evaluate transgressions are self-reported [@DePaulo1998; @DePaulo2004], they do not directly quantify the breaking of promises [@Vanberg2008; @Charness2006], or as we will discuss later, they use heterogeneous measures of social closeness [@Glaeser2000]. Our results suggest that the mere fact of considering that subjects play with a human diminishes the probability of breaking the promise. It is more likely to break the promise to the computer than to the stranger, even though that partner was not known. Also, zero social closeness is important to include because it allows us to explore the intrinsic motivation to keep the promises. If humans keep their word because it is morally correct, we would anticipate that the probability of breaking a promise was very low in the three partners and, particularly, the inference regarding the probability of breaking the promise to the computer would include zero. However, we found that, although low, the probability was greater than zero, which provides evidence that seems to contradict, at least in some degree, the hypothesis of intrinsic motivation. This means that we cannot exclude the possibility that the subjects keep promises mainly due to moral motivation because, even in the case of the computer, the subjects kept a large proportion of promises. We can, however, exclude that this is their _only_ motivation.

High social closeness allows us to explore the instrumental motivation to keep promises [@Baumgartner2009]. In our study, subjects were more likely to break the promise to the stranger than to the friend. If humans keep their word to facilitate cooperation in the future, we would anticipate that there would be less of a chance to break the promise with a partner of high social closeness (with whom they are very likely to cooperate, even after the experiment) and, in fact, the promise to friends was not broken on any occasion. The findings concerning motivations could be explained by the social norm called "conditional cooperation", which indicates that the belief that other people cooperate at high levels also induces high levels of cooperation [@Fehr2018]. Thus, the uncertainty regarding the computer’s decisions could explain the probability of breaking the promise. Similarly, the information that the subjects have regarding their friend’s behavior, even before the experiment, could explain the high levels of cooperation and keeping of promises towards the friend.

Cooperation has been studied with diverse experimental manipulations. In one study, the contribution in a “public goods game” was evaluated during several trials in a group made up of the same individuals (partners), and compared to another group of new different subjects for each trial (strangers) [@Croson1996]. In the condition of partners, the cooperation was greater because it was a stable group compared to the group of strangers. The mentioned study does not give details regarding the recruitment of participants, thus we could assume that even in the condition of partners these were individuals who did not consider themselves socially close. Another study showed an increase in cooperation in a trust game when individuals were socially close [@Glaeser2000]. In this study, subjects knew each other and the researchers carefully measured several variables regarding their social connection such as the number of friends in common and years of relationship. They found greater cooperation depending on the degree of social connection. However, the study did not control for social closeness as some individuals who arrived together were allowed to perform the task together whereas others were paired differently. This limitation allowed social closeness between partners to be heterogeneous, allowed for romantic relationships and assumed that two individuals who arrived together to class were considered close to each other. To try to homogenize social closeness between our subjects and their partners, they performed the task with a same-sex friend considered to be close, emphasizing their companion could not be a romantic or sexual partner. Bayesian analysis of the IOS scale showed evidence in favor that the degree of social closeness between 30 of our 45 subjects and their friends was the same. Although results of the two mentioned studies also found that social closeness enhance cooperation [@Croson1996; @Glaeser2000], it would be valuable to homogenize both the performed tasks and the statistical procedure used to establish similarity in effects magnitudes.

Our methodological contribution to the field is the use of Bayesian inference tools: Hierarchical Models and Cumulative Models. The Hierarchical Models, allowed us to model the effect of the experimental conditions on the individual response. These models assume that lower level observations (e.g. decisions of each individual) are nested in higher level units (e.g. individual subjects). Within-subjects designs have traditionally been analyzed with repeated measures AN(C)OVA, however, Hierarchical Models grant the advantages of naturally dealing with unbalanced data, include categorical and/or numerical predictors, explicitly incorporate individual variability, among others [@Vuorre2018; @Gelman2006; @McElreath2018; @Burkner2017].  In our case, the Hierarchical Model allowed us to capture how the experimental conditions affected the decisions of each subject and the differences between them. For example, there were obviously motivated instrumental subjects such as number 8 or 12 who are very sensitive to the identity of their investors and modify their cooperation under social closeness. At the same time, there were notably intrinsic subjects such as 9 and 10 who cooperate all the time regardless of who their partners are. The Hierarchical Model naturally includes this information for the estimation of population effects, which, if not considered, would lead to inaccurate inferences [@Burkner2017; @Gelman2006; @McElreath2018]. The Cumulative Model allowed us to capture the strength of the commitment expressed through promises. A very frequent problem that has been pointed out recently is that analyzing ordinal data with methods that assume that observations are metric can lead to serious inference errors [@Liddell2018]. The Cumulative Models assumes that the observed responses come from the categorization of a continuous latent variable [@Burkner2019]. In our study, we could identify that groups with different payment proportions, which in other studies have been called “dishonest” and “honest” [@Baumgartner2009; @Baumgartner2013], differ in the commitment they express through their promises. Considering that the tools to perform the Cumulative Models are relatively recent, in future studies, it would be valuable to also use the new alternatives for inference. Although the Hierarchical and Cumulative models are not tools exclusively for Bayesian inference, their application from this approach represents several advantages compared to frequentist statistics. Classic problems such as multiple comparisons, the decision of when to stop collecting subjects (stopping rule), or use of planned comparisons versus post hoc, are not factors that affect the Bayesian approach  [@Dienes2011]. In Bayesian inference, hypothesis testing makes formal use of probability to express the plausibility of theories, and in our case, we were able to obtain evidence ratios regarding the extent of our data support hypotheses, regardless if these were proposed _a priori_ or _post hoc_.

@@

A limitation of our study was the use of hypothetical rewards compared to real rewards. A relatively recent study on decision-making reported that there is more loss aversion when subjects have real monetary rewards compared to hypothetical in a risk task [@Xu2016]. However, other works do not report differences between the use of both types of rewards in self-control, temporal and social discount tasks [@Locey2011; @Johnson2006; @Rachlin2006]. Likewise, it can be argued that our results have theoretical congruence  [@Fehr2018] and are in the same direction as other works that use real money [@Croson1996; @Glaeser2000]. Thus, there are not many reasons to expect that other types of rewards would modify our results. Another possible limitation could be the use of multiple trials with each partner. Although the decision to use repeated measures during our design serves the purpose of reducing contaminant sources and increasing internal validity  [@Maxwell2018], the studies could benefit from only including one trial for each partner to have more clarity regarding the difference in the intrinsic and instrumental motivations. Play a one-shot with each partner avoids the possibility that multiple trials could generate the belief that the computer can also vary its behavior according to the decisions of the trustee.

@@

Subjects in our study seemed to keep their promises by a combination of _intrinsic_ and _instrumental_ motivations. They can be regarded as predominantly "nice people" because they kept their word even with investors with no social closeness (computer). However, their trustworthiness was far from perfect, since there was a small proportion of betrayals committed towards the computer. If subjects could not be sure their investor would be a "potential cooperator", dishonesty may have emerged in the form of broken promises. Framing our findings in terms of the theoretical predictions of the social norm "conditional cooperation", social closeness decreases the probability of breaking the promise and increases cooperation, and suggests the correspondence between commitment expressed through promises and subsequent behavior.
